\documentclass[review,3p,times]{elsarticle}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}

\journal{Advanced Engineering Informatics}

\newcommand{\breeze}{BREEZE}
\newcommand{\pucond}{N09\_M07\_F10}

\begin{document}
\begin{frontmatter}

\title{Training-free, physics-verified LLM recipe generation for industrial machinery condition monitoring}

\author[a]{Author One}
\author[a]{Author Two}
\affiliation[a]{organization={Department}, city={City}, country={Country}}

\begin{abstract}
Training a signal generator for each machine and data regime can be impractical
when fault examples are scarce. This paper studies a different design point:
an LLM proposes a structured signal recipe, a deterministic renderer produces
the waveform, and train-calibrated physical gates decide whether the candidate
may enter an augmentation pool. The gates are admission records, not proofs of
physical validity. Under a file-disjoint PU bearing protocol, the frozen
Phase-A v2 experiment shows that LLM recipes outperform rule recipes and
random open-loop recipes at $n_{\rm real}=5,10,25$ for both Accuracy and
Macro-F1 after the registered Holm correction (20 seeds). A separate CWRU
within-load and leave-one-load-out protocol yields 90/90 registered
LLM-comparator tests passing Holm correction (40 seeds). On Berkeley milling,
the healthy-versus-degraded binary formal test is deliberately reported as
partial: 15/18 registered comparisons pass, including all 12 comparisons
against non-structured baselines; against the rule baseline, only Accuracy at
$n_{\rm real}=2$ and both metrics at $n_{\rm real}=5$ pass, while both
metrics at $n_{\rm real}=10$ do not. Physical diagnostics are reported as
reference-relative evidence rather than a global generator ranking. We also
report six failed PU cross-condition attempts, experimental-condition and
process-metadata confounding in UMich milling, and a stopped MU-TCM inner
validation line. These results support an auditable, training-free recipe
admission workflow within the stated few-shot protocols, while defining its
current boundary for cross-condition extrapolation and confounded milling
data.
\end{abstract}

\begin{keyword}
Condition monitoring \sep Large language model \sep Synthetic data \sep
Physics-based verification \sep Few-shot diagnosis \sep Training-free generation
\end{keyword}
\end{frontmatter}

\section{Introduction}
\label{sec:intro}

Synthetic augmentation is often used when labelled fault data are scarce, but
the usual solution is to train a GAN, VAE, or diffusion model on the available
signals. The cost and data requirement of that training are a poor fit for a
small new dataset, a changing machine configuration, or a closed-source model
that cannot be updated. LLMs offer another route: they can propose a compact,
machine-conditioned recipe that a deterministic renderer turns into a signal
window. The resulting flexibility does not make the output trustworthy. A
plausible recipe can still yield a waveform with unsupported statistics,
spectral structure, or fault-frequency evidence.

\breeze\ treats this problem as admission rather than waveform repair. The LLM
proposes a recipe; the renderer has responsibility for the signal equation;
and a verifier, calibrated only on the training split, admits or rejects the
rendered candidate. Failed candidates receive a structured report for a bounded
number of resampling attempts. The verifier is not trained, it does not inspect
the test split, and it does not modify a rejected waveform. A pass is therefore
an auditable statement that a candidate met the configured, train-supported
checks, not a certificate of physical correctness or downstream generalization.

This distinction matters for evaluation. Window-level random splits can place
closely related windows from the same recording on both sides of a split, so
their accuracy should not be compared directly with results from file- or
experiment-disjoint protocols. We use only the frozen split manifests and
paired statistical families registered for each dataset. We also keep negative
experiments in the paper: six PU cross-condition attempts failed, and two
milling lines were stopped before a formal test. These are boundaries of the
method, not missing positive results.

The contribution is fourfold. First, we present a training-free
recipe--renderer--verifier workflow for condition-monitoring signals, with
clear responsibility boundaries. Second, we evaluate it under strict few-shot
protocols on PU and CWRU, with paired Holm-corrected comparisons. Third, we
report a qualified Berkeley milling result and class-conditional physical
diagnostics rather than claiming universal or state-of-the-art superiority.
Fourth, we document where the evidence fails: cross-condition morphology does
not extrapolate under the tested PU protocols, and available milling metadata
can confound wear labels.

\section{Related work and positioning}
\label{sec:related}

GANs, VAEs, and diffusion models learn a signal distribution through model
training~\cite{goodfellow,kingma,ho,timegan}. Recent bearing systems also
combine language-derived descriptions with a learned diffusion generator, as
in BearGen~\cite{beargen}. Such systems solve a different problem from
\breeze: they invest data and optimization into a learned generator. \breeze
does not claim to replace them. Its question is whether a candidate proposed
by a black-box LLM recipe can be rendered and screened without training a new
generator for the target machine.

This is also different from direct code or waveform generation. In a
recipe-as-data design, the LLM produces named physical parameters and a
deterministic renderer executes the signal model. The separation makes it
possible to audit the proposal, renderer, and verifier independently. Bearing
diagnostic tools such as envelope analysis and spectral kurtosis provide the
domain evidence used by the verifier~\cite{randall,antoni,antoni_kurtogram};
they do not turn the verifier into a complete physical simulator. Current
sideband evidence is used only when the training data support it
reliably~\cite{schoen}.

\section{Method}
\label{sec:method}

Let $\mathcal{D}_{\rm tr}=\{(x_i,y_i,m_i)\}$ denote training windows, labels,
and available machine metadata. For class $y$ and metadata $m$, an LLM proposes
a recipe $r_t$ at feedback round $t$. A deterministic renderer $R$ constructs
$\tilde{x}_t=R(r_t)$, and a verifier $V$ returns a pass indicator and an
auditable report:
\begin{equation}
  (a_t, q_t) = V(\tilde{x}_t,y,m,\mathcal{D}_{\rm tr}),\qquad
  a_t\in\{0,1\}.
\end{equation}
The candidate is admitted only when $a_t=1$. Otherwise, the next prompt
receives the structured report $q_t$ until the predeclared limit $K$ is
exhausted. No test window enters $R$, $V$, or feedback construction.

The active gates check finite format and non-degeneracy, train-supported
time-domain statistics, spectral shape, class-relevant envelope evidence,
reliability-gated current evidence where applicable, and diversity relative to
the accepted pool. Each check is calibrated on the appropriate outer training
fold. The gate family is intentionally mixed: some features require an
interval, some impose an upper or lower bound, and diversity is a
nearest-neighbour exclusion rule. This is an engineering admission criterion,
not a likelihood model or a proof that a window belongs to the real data
distribution.

\section{Experimental protocols}
\label{sec:setup}

\paragraph{PU Phase-A v2.} The primary protocol uses the PU \pucond\ file
split. It contains 3846 training windows and 962 test windows over healthy,
outer-race (OR), and inner-race (IR) classes. The renderer and verifier use
only the training partition. The frozen comparison uses balanced pools of 150
synthetic windows per class, 20 paired CNN seeds, $n_{\rm real}=5,10,25$, and
registered one-sided Wilcoxon/Holm families for LLM versus rule and random
open-loop recipes.

\paragraph{CWRU.} The frozen CWRU patch-v2 protocol uses 40 seeds,
$n_{\rm real}=5,10,25$, and $n_{\rm syn}=20$ per class. It evaluates a
within-load0 split and four leave-one-load-out folds. The CWRU verifier has its
own bearing configuration; no PU current-sideband gate is reused by default.

\paragraph{Berkeley milling.} The formal milling task is binary: healthy
means $VB<0.2$ and degraded means $VB\geq0.2$. It uses a one-time 40-seed
formal test at $n_{\rm real}=2,5,10$ with $n_{\rm syn}=20$ per class. The
registered comparison family contains LLM versus rule, noise augmentation, and
random open-loop for Accuracy and Macro-F1. Its result is reported with the
full pass/fail pattern, rather than reduced to a single aggregate claim.

\paragraph{Training-based baselines.} TimeGAN and a 1D DDPM are evaluated
under the frozen PU file split using the same downstream CNN and shot counts.
Each generator is trained only on the outer training data, once in a
full-training-fold regime and once in a few-shot regime. The implementation
checkpoints every epoch and records raw optimization trajectories, failures,
generated-pool provenance, and wall time. These results are not used in the
claims below until the predeclared 40-seed runs complete; smoke outputs are
explicitly excluded.

\section{Results}
\label{sec:results}

\subsection{PU primary condition}

The primary result is confined to the frozen PU file-split protocol. At all
three registered few-shot counts, LLM $K=3$ recipes are significantly better
than both rule and random-open-loop recipes for Accuracy and Macro-F1 after
within-family Holm correction (Table~\ref{tab:pu-phasea}). The result does not
establish zero-real-data operation: the verifier and recipe construction depend
on the outer training fold and class exemplars.

\input{generated/pu_phasea.tex}

\subsection{CWRU within-load and leave-one-load-out}

On CWRU, the frozen within-load means favour LLM recipes over the rule and
noise baselines at each shown shot count (Table~\ref{tab:cwru}). The stronger
statistical statement is the registered one: all 90 LLM-comparator tests across
the within-load and four LOLO folds pass their Holm correction. This result is
specific to the CWRU split manifests and its train-calibrated verifier; it is
not evidence that a single bearing configuration transfers without
recalibration.

\input{generated/cwru.tex}

\subsection{Berkeley milling is a qualified, partial result}

Berkeley is not reported as a passed milling benchmark. The formal result has
15/18 passing registered comparisons. All 12 LLM comparisons against the
non-structured noise and random-open-loop baselines pass. The rule comparison
is narrower: at two shots, only Accuracy passes; at five shots, both metrics
pass; and at ten shots, neither metric passes. Thus the supported claim is
that structured physical recipes outperform all non-structured baselines in
this experiment, while LLM recipes have an advantage over the rule baseline
only in the lower-shot settings.

\input{generated/berkeley.tex}

\subsection{Physical diagnostics}

Physical diagnostics compare class-conditional synthetic pools with the
matching outer-training reference. They include fault-frequency envelope
alignment for bearing faults, RMS and kurtosis Wasserstein distances, PSD-CDF
distance, band-energy error, and nearest-neighbour diversity. The main-text
PU summary in Table~\ref{tab:physics} is generated from fixed 150-window/class
pools; CWRU and Berkeley tables, complete class-wise values, source hashes,
and availability records are retained with the experiment artifacts.

These diagnostics are not a universal ranking. For example, a single
fault-frequency alignment diagnostic can favour a rule pool even when other
reference-relative diagnostics and the downstream result differ. A matching
frozen CWRU random-open-loop pool does not exist, and the table records that
absence instead of substituting a different pool. TimeGAN and DDPM pools will
be added only after their non-smoke formal runs finish.

\input{generated/physics.tex}

\section{Negative results and scope boundaries}
\label{sec:boundaries}

The scope boundaries are part of the evidence. Six preregistered PU LOCO
attempts failed. The sequence progressed from frequency mismatch through
insufficient frequency correction and morphology extrapolation failures to a
source-only CSCoh diagnostic that was non-discriminative for the target
classes. We therefore do not claim cross-speed or cross-load morphology
transfer.

For UMich milling, the correct conclusion is not that the signals are
generically unlearnable. The available experimental-condition and process
metadata are confounded with wear labels under the current split, and the
confounding cannot be disentangled from wear without changing the protocol or
acquiring additional metadata. For MU-TCM v3, the inner-validation gate passes
only 2/6 core comparisons. No preregistration or formal held-out test was run.
Neither line is used as a positive result.

\section{Discussion}
\label{sec:discussion}

The results support a limited practical statement. A recipe-based,
training-free admission workflow can improve few-shot diagnosis under the
specified PU and CWRU protocols, and its decisions can be inspected through
renderer inputs and gate reports. It is not a substitute for a learned
generator in every data regime. The current trained-baseline study is designed
to quantify that trade-off directly through matched downstream performance,
data requirement, and measured training cost; no cost superiority claim is
made before that study completes.

The Berkeley result illustrates why statistical detail matters. An average
accuracy alone would hide the fact that rule and LLM converge at ten shots,
and that the aggregate result is only partial. Similarly, physical fidelity is
not synonymous with downstream utility. The verifier gives an auditable
admission decision, but a candidate can satisfy the measured gates without
capturing every mechanism relevant to a classifier or to a field deployment.

\section{Conclusion}
\label{sec:conclusion}

\breeze\ separates LLM recipe proposal, deterministic signal rendering, and
train-calibrated physical admission. On frozen file- and load-disjoint
protocols, it yields positive few-shot results on PU and CWRU, a qualified
Berkeley result, and explicit negative evidence for cross-condition and
confounded milling settings. The central contribution is therefore an
auditable workflow with stated limits, not a claim of universal generator
superiority or zero-data fault diagnosis.

\section*{Reproducibility and data availability}

All numerical tables in this manuscript are generated from frozen CSV files by
the script at \url{scripts/build_paper_tables.py}. The source ledger is
\texttt{analysis/evidence\_ledger.md}; physical-metric provenance and
availability records are in
\texttt{analysis/section\_C\_physics\_freeze\_report.md}. The PU, CWRU, and
Berkeley result directories preserve their respective split manifests and
registered test families. The Paderborn and CWRU data are public benchmark
data. Berkeley milling preprocessing, aggregate outputs, and code can be
shared subject to the originating data-use terms.

\section*{Declaration of Generative AI and AI-Assisted Technologies}

During preparation of this work, the authors used OpenAI Codex to assist with
code inspection, experiment scripting, result-table assembly, and manuscript
drafting. The authors reviewed and verified the content and take responsibility
for the final manuscript.

\bibliographystyle{elsarticle-num}
\begin{thebibliography}{00}

\bibitem{goodfellow}
I. Goodfellow, J. Pouget-Abadie, M. Mirza, B. Xu, D. Warde-Farley,
S. Ozair, A. Courville, Y. Bengio, Generative adversarial nets, in:
Advances in Neural Information Processing Systems, 2014.

\bibitem{kingma}
D. P. Kingma, M. Welling, Auto-encoding variational Bayes, in:
International Conference on Learning Representations, 2014.

\bibitem{ho}
J. Ho, A. Jain, P. Abbeel, Denoising diffusion probabilistic models, in:
Advances in Neural Information Processing Systems, 2020.

\bibitem{timegan}
J. Yoon, D. Jarrett, M. van der Schaar, Time-series generative adversarial
networks, in: Advances in Neural Information Processing Systems, 2019.

\bibitem{beargen}
J. Lee, H. Jeon, U. Kim, M. Kim, BearGen: LLM-guided signal generation
framework for bearing fault diagnosis, Advanced Engineering Informatics 71
(2026) 104400. \url{https://doi.org/10.1016/j.aei.2026.104400}.

\bibitem{randall}
R. B. Randall, J. Antoni, Rolling element bearing diagnostics--a tutorial,
Mechanical Systems and Signal Processing 25 (2011) 485--520.

\bibitem{antoni}
J. Antoni, The spectral kurtosis: a useful tool for characterising
non-stationary signals, Mechanical Systems and Signal Processing 20 (2006)
308--331.

\bibitem{antoni_kurtogram}
J. Antoni, Fast computation of the kurtogram for the detection of transient
faults, Mechanical Systems and Signal Processing 21 (2007) 108--124.

\bibitem{schoen}
R. R. Schoen, T. G. Habetler, F. Kamran, R. G. Bartheld, Motor bearing damage
detection using stator current monitoring, IEEE Transactions on Industry
Applications 31 (1995) 127--138.

\end{thebibliography}
\end{document}
